---
title: "final_wrangle"
author: "Riku Turkki"
date: "3/6/2017"
output: html_document
---
contact: [riku.turkki@helsinki.fi](mailto:riku.turkki@helsinki.fi)

### Import and data wrangling

I start by reading the *student-mat* data set from a local csv file.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)

(data <- read_delim("student-mat.csv",";"))

```

As can be seen above, the whole data set contains 33 variables and 395 observations. To continue, I select 11 variables and only those rows that describes female students. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}

(data <- data %>%
   filter(sex == "F") %>%
   dplyr::select(age, Medu, Fedu, studytime, goout, Walc, Dalc, G3)) 
  
```

Finally, my data set comprise 208 students with following 8 variables:

* age - student's age (numeric: from 15 to 22) 

* Medu - mother's education (numeric: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education or 4 - 
higher education) 

* Fedu - father's education (numeric: 0 - none, 1 - primary education (4th grade), 2 - 5th to 9th grade, 3 - secondary education or 4 - 
higher education) 

* studytime - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours) 

* goout - going out with friends (numeric: from 1 - very low to 5 - very high) 

* Dalc - workday alcohol consumption (numeric: from 1 - very low to 5 - very high) 

* Walc - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high) 

* G3 - final grade (numeric: from 0 to 20, output target) 

### Saving the wrangled data. 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
saveRDS(data, "my_data.rds")
```
